Discrepancy, Coresets, and Sketches in Machine Learning

### 2019-02-02 komlos.tex
edits to intro

An important application of our generic coreset result is Kernel density estimation.
Kernel density estimation is a popular tool in data analysis aimed to estimate a continuous distribution with a finite set of points. Among other applications, this tool is used for outlier detection \cite{schubert2014generalized}, regression \cite{fan2018local}, and clustering \cite{rinaldo2010generalized}. A thorough survey could be found in \cite{silverman2018density}. Given a set of $n$ data points $x_1,\ldots,x_n$ and a query $q$, the density estimate of $q$ is the average $\frac{1}{n} \sum_i K(x_i,q)$ of the kernel evaluations with the data points. The task of finding a coreset for kernel density estimates consists of obtaining a set of points $z_1,\ldots,z_m$ with weights $w_1,\ldots,w_m$ for some $m \ll n$ such that for any query $q$ it holds that 
$\frac{1}{n} \sum_i K(x_i,q) = \frac{1}{m} \sum_i w_i K(z_i,q) \pm \eps$. For this task, the state-of-the-art is given by \cite{DBLP:journals/corr/abs-1802-01751}, achieving $m=\sqrt{d\log(1/\eps)}/\eps$ where $d$ is the dimension of the original data points. Their result holds for any Lipchitz bounded positive semi-definite kernels. The result is constructive though it is polynomial rather than (quasi-)linear in the data size. The authors give an almost matching lower bound of $\sqrt{d}/\eps$ and pose an open question for closing the gap between the bounds. Additionally, they provide a simpler algorithm based on the Frank-Wolf method that achieves a $1/\eps^2$ sized coreset. It is in fact known (see \cite{lopez2015towards}, Theorem 1) that a uniform random sample of $\log(1/\delta)/\eps^2$ points gives a coreset w.p. $1-\delta$ for some kernel types. 
In this paper we resolve the open question by \cite{DBLP:journals/corr/abs-1802-01751} and prove that the coreset complexity of Gaussian kernel density is indeed $O(\sqrt{d}/\eps)$, matching the lower bound. 
In fact, we show that this is the coreset complexity for any analytic function of the squared distance $f(x,q) = g(\|x-q\|^2)$.
Moreover, we provide a very simple and deterministic algorithm for constructing coresets of size $1/\eps^2$ for any positive kernel. Although the worst case lower bound is $1/\eps^2$, we conjecture that in realistic scenarios, the deterministic algorithm should outperform random sampling as it explicitly takes advantage of negative correlations.

Another application of our coreset results applies to logistic regression. Logistic Regression is an extremely popular technique for classification. We are given a set of $n$ pairs $x_i,y_i$ with $x_i \in \R^d$ and $y_i \in \{-1,1\}$. The objective is to find a regressor $q \in \R^d$ minimizing the logistic loss defined as $\E_i \log(1+\exp(x_i^\top q y_i))$. A coreset for logistic regression is a set of $m \ll n$ pairs $z_i, v_i$ in $\R^d, \{-1,1\}$ correspondingly along with weights $w_i$, with the property that for any regressor $q$, either $\E_i \log(1+\exp(x_i^\top q y_i)) = \E_i \log(1+\exp(z_i^\top q v_i)) \pm \eps$ or $\E_i \log(1+\exp(x_i^\top q y_i)) = \E_i \log(1+\exp(z_i^\top q v_i)) (1 \pm \eps)$. A recent paper \cite{DBLP:journals/corr/abs-1805-08571} provides a coreset with a multiplicative guarantee that is based on an average sensitivity property of the dataset. They provide a lower bound for the size of a multiplicative coreset showing that in general it is not possible to achieve $m \ll n$. The coreset they build is of cardinality $m \approx \mu\sqrt{nd^3}/\eps^2$ where $\mu \geq 1$ is a complexity measure of the dataset.  \cite{tolochinsky2018coresets} gives a generic multiplicative coreset construction for any monotonic function with $\ell_2^2$ regularization. The dependence they get is $\tilde O(d/\eps^2)$ ignoring logarithmic factors.  Our results prove that for an additive guarantee, it is possible to obtain a coreset of size $O(\sqrt{d}/\eps)$.

>>>>

We show that our result applies to any analytic function of the dot product. 
These include Logistic Regression $F(q) =  \sum_i \log(1+\exp(\ip{y_ix_i,q}))$, Covariance or matrix approximation  $F(q) =  \sum_i \ip{x_i,q}^2$, sigmoid activation loss $F(q) =  \sum_i 1/(1+\exp(\ip{y_ix_i,q}))$, linear regression $F(q) =  \sum_i (\ip{y_ix_i,q} -y_i)^2$ and many others. For all the aforementioned  $x,q \in R^d$ and $y_i \in \{-1,-1\}$.
By bounding the class discrepancy of all such functions we obtain coresets of size $O(\sqrt{d}/\eps)$ for all of them.

We note that while we obtain universal {\it additive guarantee} it is often much harder to get a  {\it multiplicative guarantee}
For Logistic regression, for example, a recent paper \cite{DBLP:journals/corr/abs-1805-08571} provides a coreset with a  {\it multiplicative guarantee} that is based on an average sensitivity property of the dataset. 
They provide a lower bound for the size of a multiplicative error coreset.
They show that in general it is not possible to achieve $m \ll n$. 
The coreset they build is of cardinality $m \approx \mu\sqrt{nd^3}/\eps^2$ where $\mu \geq 1$ is a complexity measure of the dataset.  \cite{tolochinsky2018coresets} gives a generic multiplicative coreset construction for any monotonic function with $\ell_2^2$ regularization. 
The dependence they get is $\tilde O(d/\eps^2)$ ignoring logarithmic factors.

We show that our result also applies to any analytic function of the squared distance.
A prime example of that is Gaussian kernel density estimation.
Kernel density estimation is a popular tool in data analysis aimed to estimate a continuous distribution with a finite set of points. 
Among other applications, this tool is used for outlier detection \cite{schubert2014generalized}, regression \cite{fan2018local}, and clustering \cite{rinaldo2010generalized}. 
A thorough survey could be found in \cite{silverman2018density}. 
Given a set of $n$ data points $\{x_1,\ldots,x_n\}$ and a query $q$, the Gaussian density estimate at point $q$ is given by $\sum_i K(x_i,q) = \sum_i e^{-\|x-q\|^2}$. 
Obtaining the smallest possible coreset for this problem has been open for several years.
The state-of-the-art is given by \cite{DBLP:journals/corr/abs-1802-01751} (see references within). 
They achieve coresets of size $O(\sqrt{d\log(1/\eps)}/\eps)$ where $d$ is the dimension of the original data points. 
Their result holds for any Lipchitz bounded positive semi-definite kernels. 
The result is constructive though it is polynomial rather than (quasi-)linear in the data size. 
The authors give an almost matching lower bound of $\sqrt{d}/\eps$ and pose an open question for closing the gap between the bounds. 
Additionally, they provide an algorithm based on the Frank-Wolf method that achieves a $1/\eps^2$ sized coreset. 
It is, in fact, known (see \cite{lopez2015towards}, Theorem 1) that a uniform random sample of $\log(1/\delta)/\eps^2$ points gives a coreset w.p. $1-\delta$ for some kernel types. 
In this paper we resolve the open question by \cite{DBLP:journals/corr/abs-1802-01751} and prove that the coreset complexity of Gaussian kernel density is indeed $O(\sqrt{d}/\eps)$, matching the lower bound. 
In fact, we show that this is the coreset complexity for any analytic function of the squared distance $f(x,q) = f(\|x-q\|^2)$.

We also provide another independent result; 
a very simple and deterministic algorithm for constructing coresets of size $O(1/\eps^2)$ for any positive semi-definite kernel. 
Although the worst case lower bound is $O(1/\eps^2)$ which matches the coreset achieved by sampling, we conjecture that in realistic scenarios, the deterministic algorithm will outperform random sampling as it explicitly takes advantage of negative correlations.
